---
title: "2025_BCT_63_Assignment2"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
rm(list = ls())
```

Question 1

Dataset/Structure: Vectors in Base R

Concepts: NA values, Vectorized Comparisons, Logical Indexing

Write an R script that performs the following steps:

1\. Create a numeric vector observation_scores containing the values:
12, NA, 45, 8, NA, 67, 23, 0, -5, NA, 89.

```{r}
observation_scores = c(12, NA, 45, 8, NA, 67, 23, 0, -5, NA, 89)
```

#2. Write a single logical expression to identify indices that are not
missing AND are strictly greater than 15.

```{r}
!is.na(observation_scores) & observation_scores > 15
```

#3. Replace all NA values in observation_scores with the calculated
median of the non-missing elements.

```{r}
sum(observation_scores, na.rm=T)/11 
observation_scores[is.na(observation_scores)]=21.72727
```

#4. Print the final modified vector along with a text message stating
its new mean.

```{r}
print(mean(observation_scores))
```

Question 2

Dataset/Structure: Special Values (Inf, -Inf, NaN)

Concepts: Indeterminate Forms, Boundary Arithmetic, Testing Functions

1\. Write an R code snippet that computes and prints the results of the
following four mathematical operations:

○ $0 / 0$

```{r}
nan_ex <- 0/0
```

#○ $45 / 0$

```{r}
inf_ex <- 45/0
```

#○ $-12 / 0$

```{r}
neg_inf_ex <- -12/0
```

#○ $\log(0)$

```{r}
log_ex <-log(0)
```

#2. Use the built-in functions is.na(), is.nan(), and is.infinite() to
test each of the four results. #3. Display the results in a structured
tabular output or clear console logs, and write a brief comment
explaining why is.na() returns TRUE for NaN values in R.

```{r}
is.na(nan_ex) #is.na(nan_ex) returns True because in R, the is.na() function is designed to detect both missing values (NA) and undefined numeric results (NaN).
is.nan(nan_ex) #is.nan(nan_ex) returns TRUE only if the variable nan_ex contains the specific numeric value NaN ("Not a Number").
is.infinite(nan_ex)#is.infinite(nan_ex) returns FALSE because NaN ("Not a Number") is mathematically distinct from infinity.

is.na(inf_ex) #is.na(inf_ex) returns FALSE because infinity (Inf) is a defined, valid numeric value in R, not a missing or undefined result.
is.nan(inf_ex) #is.nan(inf_ex) returns FALSE because infinity (Inf) is a defined numeric limit, whereas NaN stands for "Not a Number" representing an undefined result.
is.infinite(inf_ex) #is.infinite(inf_ex) returns TRUE because the variable inf_ex contains the value Inf (infinity), which is exactly what this function is designed to detect.

is.na(neg_inf_ex) #is.na(neg_inf_ex) returns FALSE because negative infinity (-Inf) is a defined, valid numeric value in R, not a missing or undefined result.
is.nan(neg_inf_ex) #is.nan(neg_inf_ex) returns FALSE because negative infinity (-Inf) is a defined numeric limit, whereas NaN stands for "Not a Number" representing an undefined result.
is.infinite(neg_inf_ex) #is.infinite(neg_inf_ex) returns TRUE because the variable neg_inf_ex contains the value -Inf (negative infinity), which this function is specifically designed to detect.

is.na(log_ex) #is.na(log_ex) returns FALSE because the variable log_ex contains a defined numeric value, not a missing (NA) or undefined (NaN) result.
is.nan(log_ex) #is.nan(log_ex) returns FALSE because the value stored in log_ex is not "Not a Number" (NaN).
is.infinite(log_ex) #is.infinite(log_ex) returns TRUE because the variable log_ex contains the value -Inf (negative infinity).
```

Question 3

Dataset/Structure: Base R Dataset airquality

Concepts: Data Inspection, Missing Value Profiling, Column-wise NA
Counting

Load the built-in airquality dataset and perform the following analysis:

1\. Inspect the total dimensions (rows and columns) of airquality.

```{r}
library(airqualityES) 
dim(airquality)
```

#2. Calculate the total number of NA values present in the entire
dataset.

```{r}
sum(is.na(airquality))
```

#3. Calculate the number of NA values present in each individual column
(Ozone, Solar.R, Wind, Temp, Month, Day).

```{r}
sum(is.na(airquality$Ozone))
sum(is.na(airquality$Solar.R)) 
sum(is.na(airquality$Wind))
sum(is.na(airquality$Temp)) 
sum(is.na(airquality$Month))
sum(is.na(airquality$Day))
```

#4. Calculate the mean of the Ozone column without removing NA values,
observe the output, and then calculate the correct mean by explicitly
stripping NA values using na.rm = TRUE.

```{r}
mean(airquality) 
mean(airquality$Ozone, na.rm=TRUE) 
```

Question 4

Dataset/Structure: Base R Dataset airquality

Concepts: complete.cases(), na.omit(), Dataset Subsetting

1\. Using the airquality dataset, extract all rows that contain at least
one missing value (NA) using the !complete.cases() function and store
them in a new data frame called air_incomplete. Print the first 6 rows
of air_incomplete.

```{r}
dim(airquality)

!complete.cases(airquality)
air_incomplete <- airquality[!complete.cases(airquality),]
head(air_incomplete)
```

2\. Create a fully clean dataset called air_clean by removing all rows
containing missing values using na.omit().

```{r}
air_clean <- na.omit(airquality)
```

3\. Verify that the sum of the row counts of air_incomplete and
air_clean exactly equals the total number of rows in the original
airquality dataset.

```{r}
total_original <- nrow(airquality)
total_split <- nrow(air_incomplete) + nrow(air_clean)

# Check equality
total_original == total_split   
```

Section B: Subsetting, Deduplication & Data Manipulation (Total Marks:
30 Marks \| 10 Marks Each)

Question 5

Dataset/Structure: Base R Dataset mtcars

Concepts: Subsetting using Bracket Notation [row, col], Logical
Conditions, Indexing Load the built-in\
dataset and write an R script to extract the following subsets using
bracket indexing [ , ]:

1\. Extract rows 5 through 15 and columns 1 through 4 (mpg, cyl, disp,
hp).

```{r}
library(car)
head(mtcars)
mtcars[5:15,1:4]
```

2\. Extract all cars (rows) that have a gas mileage (mpg) greater than
20.0 AND possess more than 100 horsepower (hp).

```{r}
mpd_hp <- subset(mtcars, mpg > 20.0 & hp > 100)
```

3\. Extract only the mpg and wt columns for vehicles with an 8-cylinder
engine (cyl == 8).

```{r}
mpg_wt_cyl <- subset(mtcars, cyl == 8, select = c(mpg, wt))
```

4\. Calculate the average weight (wt) of cars that have a automatic
transmission (am == 0) versus manual transmission (am == 1).

```{r}
avg_wt_by_am <- aggregate(wt ~ am, data = mtcars, FUN = mean)
```

Question 6 Dataset/Structure: Base R Dataset ToothGrowth

Concepts: subset() Function, Multi-Condition Filtering, Column Selection
Load the ToothGrowth dataset (which measures tooth growth in guinea pigs
based on Vitamin C delivery method and dosage) and perform the following
tasks:

1\. Display the structure (str()) and summary statistics (summary()) of
ToothGrowth.

```{r}
head(ToothGrowth)
str(ToothGrowth)
summary(ToothGrowth)
```

2\. Use the subset() function to filter observations where the
supplement type (supp) is equal to "VC" (Ascorbic Acid) AND the dose
(dose) is strictly greater than or equal to 1.0.

```{r}
sup_dose <- subset(ToothGrowth, supp == "VC" & dose >= 1.0)
```

3\. Use the subset() function to extract only the tooth length (len)
column for animals that received the "OJ" (Orange Juice) supplement at a
dose of 0.5.

```{r}
tooth_len <- subset(ToothGrowth, supp == "OJ" & dose == 0.5, select = len)
```

4\. Calculate the mean tooth length for the "VC" group versus the "OJ"
group at a 2.0 dose level using subset() and mean().

```{r}
mean_oj_2 <- mean(subset(ToothGrowth, supp == "OJ" & dose == 2.0)$len)

mean_vc_2 <- mean(subset(ToothGrowth, supp == "VC" & dose == 2.0)$len)
```

Question 7

Dataset/Structure: Base R Dataset chickwts / Artificially Duplicated
Frame Concepts: Dataset Duplication, Row Stacking rbind(), unique(),
duplicated()

1\. Load the built-in chickwts dataset and display its original
dimensions using dim().

```{r}
dim(chickwts)
```

2.  Create an artificially duplicated data frame named chick_duplicated
    by binding the first 40 rows of chickwts onto the original dataset
    using rbind().

```{r}
chick_duplicated = rbind.data.frame(chickwts, chickwts[1:40,])
```

3.  Display the dimensions of chick_duplicated and calculate the exact
    number of duplicated rows present using
    sum(duplicated(chick_duplicated)).

```{r}
dim(chick_duplicated)
sum(duplicated(chick_duplicated))
```

4.  Create a clean data frame named chick_restored by stripping out all
    duplicate rows using the unique() function.

```{r}
chick_restored = unique(chick_duplicated)
```

5\. Use an if-else block to programmatically check and print whether
dim(chick_restored) matches the dimensions of the original chickwts
dataset.

```{r}
if (dim(chick_restored)[1] == dim(chickwts)[1] && 
    dim(chick_restored)[2] == dim(chickwts)[2]) {
  print("The sizes match")
}
```

Section C: Feature Engineering & Data Restructuring (Total Marks: 25
Marks)

Question 8 (10 Marks) Dataset/Structure: Base R Dataset iris Concepts:
Derived Variables, Computed Ratios, Data Frame Column Creation Load the
built-in iris dataset and complete the following feature engineering
pipeline:

1\. Print the column names of iris using colnames().

```{r}
colnames(iris)
```

2\. Create a new derived column in iris named Petal.Ratio calculated as:
$$\text{Petal.Ratio} = \frac{\text{Petal.Length}}{\text{Petal.Width}}$$

```{r}
iris$Petal.Ratio = iris$Petal.Length/iris$Petal.Width
```

3.  Create a second derived column named Sepal.Ratio calculated as:
    $$\text{Sepal.Ratio} = \frac{\text{Sepal.Length}}{\text{Sepal.Width}}$$

```{r}
iris$Sepal.Ratio = iris$Sepal.Length/iris$Sepal.Width
```

4.  Extract all observations where Petal.Ratio \> 3.0 AND Species ==
    "virginica".

```{r}
iris[iris$Petal.Ratio>3.0 & iris$Species=="virginica",]
```

5\. Display the first 10 rows of the updated iris dataset showing only
the columns: Species, Petal.Ratio, and Sepal.Ratio.

```{r}
iris[1:10, 6:7]
```

Question 9 (15 Marks) Dataset/Structure: Package car Dataset Davis (or
base equivalent matrix mapping) Concepts: Empty Matrix/Frame
Pre-allocation, Dimension Mapping, Data Transfer, Reshaping Write an R
script that manually restructures data frame architectures without using
third-party auto-converters:

1.  Load the Davis dataset from the car package (or construct a raw
    sample frame if car is unavailable). Display its first 6 rows using
    head().

```{r}
library(car)
head(Davis)
```

2\. Programmatically determine the number of rows ($N$) and columns
($M$) of the input dataset using dim().

```{r}
dim(Davis)
```

3\. Pre-allocate an empty target data frame named restructured_output
initialized with an empty matrix of dimensions $N \times M$ using:
data.frame(matrix(nrow = N, ncol = M))

```{r}
N <- nrow(Davis)
M <- 5
restructured_output = data.frame(matrix(nrow = N, ncol = M))
```

4\. Assign custom, clean column names to restructured_output:
"Subject_Gender", "Measured_Weight", "Measured_Height",
"Reported_Weight", "Reported_Height".

```{r}
colnames(restructured_output) <- c("Subject_Gender", "Measured_Weight", "Measured_Height","Reported_Weight", "Reported_Height")
```

5\. Systematically transfer data from the original dataset columns into
the newly named columns of restructured_output.

```{r}
restructured_output$Subject_Gender   <- Davis$gender
restructured_output$Measured_Weight  <- Davis$measured_weight
restructured_output$Measured_Height  <- Davis$measured_height
restructured_output$Reported_Weight  <- Davis$reported_weight
restructured_output$Reported_Height  <- Davis$reported_height
```

6\. Add a computed column Weight_Discrepancy calculated as
Measured_Weight - Reported_Weight.

```{r}
restructured_output$Weight_Discrepancy <- restructured_output$Measured_Weight - restructured_output$Reported_Weight
```

7\. Omit all NA rows from restructured_output using na.omit() and
display the summary() of the final cleaned data frame.

```{r}
final_clean_data <- na.omit(restructured_output)
summary(final_clean_data)
```

Section D: Aggregation & Summary Pipelines with dplyr (Total Marks: 25
Marks)

Question 10 (10 Marks) Dataset/Structure: Base R Dataset mtcars
Concepts: library("dplyr"), Data Aggregation, summarize()

1\. Install (if required) and load the dplyr package using
library("dplyr").

```{r}
library(dplyr)
```

2\. Using the summarize() function from dplyr on the built-in mtcars
dataset, calculate a baseline summary report containing: ○ The mean of
mpg (miles per gallon) ○ The mean of hp (horsepower) ○ The standard
deviation (sd) of mpg ○ The standard deviation (sd) of hp

```{r}
summary(mtcars)
```

3. Store the output in a data frame named mtcars_overall_summary and
print it with formatted column names.

```{r}
mtcars_overall_summary  = summary(mtcars)
print(colnames(mtcars_overall_summary ))
```

Question 11 (15 Marks) Dataset/Structure: Base R Dataset iris Concepts:
Multi-column Summarization, Group-wise Analysis, dplyr Pipeline Write an
R script using dplyr methods to perform data summary tasks:

1\. Using the iris dataset, apply summarize() to compute: ○
Mean_Petal_Length = mean(Petal.Length) ○ SD_Petal_Length =
sd(Petal.Length) ○ Mean_Sepal_Length = mean(Sepal.Length) ○
SD_Sepal_Length = sd(Sepal.Length)

```{r}
summary(iris)
Mean_Petal_Length = mean(iris$Petal.Length)
SD_Petal_Length = sd(iris$Petal.Length)
Mean_Sepal_Length = mean(iris$Sepal.Length)
SD_Sepal_Length = sd(iris$Sepal.Length)
```

2\. Group the iris dataset by Species and calculate the mean and
standard deviation of Petal.Length for each species separately.

```{r}
iris_summary <- iris %>%
  group_by(Species) %>%
  summarise(
    Mean_Petal_Length = mean(Petal.Length),
    SD_Petal_Length = sd(Petal.Length)
  )

```

3\. Identify which species exhibits the highest average Petal.Length and
print a formatted conclusion sentence displaying the result.

```{r}
result <- iris %>%
  group_by(Species) %>%
  summarise(Mean_Petal_Length = mean(Petal.Length)) %>%
  slice_max(Mean_Petal_Length, n = 1)
```
