---
title: "2025_BCT_63_Assignment2"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
rm(list = ls())

```

Question 1

Dataset/Structure: Vectors in Base R

Concepts: NA values, Vectorized Comparisons, Logical Indexing

Write an R script that performs the following steps:

1\. Create a numeric vector observation_scores containing the values:
12, NA, 45, 8, NA, 67, 23, 0, -5, NA, 89.

```{r}
observation_scores = c(12, NA, 45, 8, NA, 67, 23, 0, -5, NA, 89)
```

#2. Write a single logical expression to identify indices that are not
missing AND are strictly greater than 15.

```{r}
!is.na(observation_scores) & observation_scores > 15
```

#3. Replace all NA values in observation_scores with the calculated
median of the non-missing elements.

```{r}
sum(observation_scores, na.rm=T)/11 
observation_scores[is.na(observation_scores)]=21.72727
```

#4. Print the final modified vector along with a text message stating
its new mean.

```{r}
print(mean(observation_scores))
```

Question 2

Dataset/Structure: Special Values (Inf, -Inf, NaN)

Concepts: Indeterminate Forms, Boundary Arithmetic, Testing Functions

1\. Write an R code snippet that computes and prints the results of the
following four mathematical operations:

○ $0 / 0$

```{r}
nan_ex <- 0/0
```

#○ $45 / 0$

```{r}
inf_ex <- 45/0
```

#○ $-12 / 0$

```{r}
neg_inf_ex <- -12/0
```

#○ $\log(0)$

```{r}
log_ex <-log(0)
```

#2. Use the built-in functions is.na(), is.nan(), and is.infinite() to
test each of the four results. #3. Display the results in a structured
tabular output or clear console logs, and write a brief comment
explaining why is.na() returns TRUE for NaN values in R.

```{r}
is.na(nan_ex) 
is.nan(nan_ex) 
is.infinite(nan_ex)

is.na(inf_ex) 
is.nan(inf_ex) 
is.infinite(inf_ex)

is.na(neg_inf_ex) 
is.nan(neg_in_fex) 
is.infinite(neg_in_fex)

is.na(log_ex) 
is.nan(log_ex) 
is.infinite(log_ex)
```

Question 3

Dataset/Structure: Base R Dataset airquality

Concepts: Data Inspection, Missing Value Profiling, Column-wise NA
Counting

Load the built-in airquality dataset and perform the following analysis:

1\. Inspect the total dimensions (rows and columns) of airquality.

```{r}
library(airqualityES) 
airquality <- airqualityES 
dim(airquality)
```

#2. Calculate the total number of NA values present in the entire
dataset.

```{r}
sum(is.na(airquality))
```

#3. Calculate the number of NA values present in each individual column
(Ozone, Solar.R, Wind, Temp, Month, Day).

```{r}
sum(is.na(airquality$Ozone))
sum(is.na(airquality$Solar.R)) 
sum(is.na(airquality$Wind))
sum(is.na(airquality$Temp)) 
sum(is.na(airquality$Month))
sum(is.na(airquality$Day))
```

#4. Calculate the mean of the Ozone column without removing NA values,
observe the output, and then calculate the correct mean by explicitly
stripping NA values using na.rm = TRUE.

```{r}
mean(airquality) 
mean(airquality$Ozone, na.rm=TRUE) 
```

Question 4

Dataset/Structure: Base R Dataset airquality

Concepts: complete.cases(), na.omit(), Dataset Subsetting

1\. Using the airquality dataset, extract all rows that contain at least
one missing value (NA) using the !complete.cases() function and store
them in a new data frame called air_incomplete. Print the first 6 rows
of air_incomplete.

```{r}
dim(airquality)

!complete.cases(airquality)
air_incomplete <- airquality[!complete.cases(airquality),]
head(air_incomplete)
```

2\. Create a fully clean dataset called air_clean by removing all rows
containing missing values using na.omit().

```{r}
air_clean <- na.omit(airquality)
```

3\. Verify that the sum of the row counts of air_incomplete and
air_clean exactly equals the total number of rows in the original
airquality dataset.

```{r}
row(air_incomplete + air_clean)
row(airquality)
```

Section B: Subsetting, Deduplication & Data Manipulation (Total Marks:
30 Marks \| 10 Marks Each) Question 5

Dataset/Structure: Base R Dataset mtcars

Concepts: Subsetting using Bracket Notation [row, col], Logical
Conditions, Indexing Load the built-in mtcars dataset and write an R
script to extract the following subsets using bracket indexing [ , ]:

1\. Extract rows 5 through 15 and columns 1 through 4 (mpg, cyl, disp,
hp).

```{r}
library(car)
```

2\. Extract all cars (rows) that have a gas mileage (mpg) greater than
20.0 AND possess more than 100 horsepower (hp).

3\. Extract only the mpg and wt columns for vehicles with an 8-cylinder
engine (cyl == 8).

4\. Calculate the average weight (wt) of cars that have a automatic
transmission (am == 0) versus manual transmission (am == 1).
