---
title: "63_Assignment5_Wrangling"
output: html_document
---

```{r}
rm(list = ls())
```

ASSIGNMENT 5: MODULE BREAKDOWN (50 MARKS)

● Task 1 Data Selection, Filtering & Pipe Transformation (dplyr) (10 Marks)

● Task 2 Feature Engineering, Conditional Mutation & Aggregation (10 Marks)

● Task 3 NA Imputation & Reshaping (tidyr: pivot_longer / wider) (10 Marks)

● Task 4 Relational Database Joins (Inner, Left, Right, Full) (10 Marks)

● Task 5 Layered Visual Analytics with ggplot2 (10 Marks)

Task 1: Data Selection, Slicing & Logical Filtering (dplyr) Objective: Master column selection, exclusion, row slicing, and multi-condition filtering on the iris dataset.

Instructions:

1\. Load library(dplyr) and inspect the built-in iris dataset using head(), str(), and summary().

```{r}
library(dplyr)
head(iris)
str(iris)
summary(iris)
```

2\. Use the native pipe operator (\|\>) with select() to display: ○ Only the columns from Sepal.Length through Petal.Length using the colon (:) operator.

```{r}
iris |> #|> is a pipe operator used to chain different functions together
  dplyr::select("Sepal.Width","Petal.Length")
```

○ Only the columns Sepal.Length, Sepal.Width, and Species.

```{r}
iris |> 
  dplyr::select("Sepal.Width","Petal.Length", "Species")
```

3\. Demonstrate column exclusion by selecting all columns except Sepal.Width using -Sepal.Width.

```{r}
iris |> 
  dplyr::select(-Sepal.Width)
```

4\. Contrast dplyr::select() with Base R bracket indexing by extracting rows 20 to 45 and columns 1 and 2 using iris[20:45, c(1, 2)].

```{r}
iris |> 
  slice(20:45) |>
  dplyr::select(,c(1, 2))
```

5\. Filter the dataset to extract: ○ Observations where Sepal.Length \> 6.0. ○ Observations where the species is either "virginica" OR "versicolor" using the logical OR operator (\|). 6. Sort the entire dataset by Sepal.Length in descending order using arrange(desc(...)) and display the first 6 rows.
