---
title: "63_Assignment5_Wrangling"
output: html_document
---

```{r}
rm(list = ls())
```

ASSIGNMENT 5: MODULE BREAKDOWN (50 MARKS)

● Task 1 Data Selection, Filtering & Pipe Transformation (dplyr) (10 Marks)

● Task 2 Feature Engineering, Conditional Mutation & Aggregation (10 Marks)

● Task 3 NA Imputation & Reshaping (tidyr: pivot_longer / wider) (10 Marks)

● Task 4 Relational Database Joins (Inner, Left, Right, Full) (10 Marks)

● Task 5 Layered Visual Analytics with ggplot2 (10 Marks)

Task 1: Data Selection, Slicing & Logical Filtering (dplyr) Objective: Master column selection, exclusion, row slicing, and multi-condition filtering on the iris dataset.

Instructions:

1\. Load library(dplyr) and inspect the built-in iris dataset using head(), str(), and summary().

```{r}
library(dplyr)
head(iris)
str(iris)
summary(iris)
```

2\. Use the native pipe operator (\|\>) with select() to display: ○ Only the columns from Sepal.Length through Petal.Length using the colon (:) operator.

```{r}
iris |> #|> is a pipe operator used to chain different functions together
  dplyr::select("Sepal.Width","Petal.Length")
```

○ Only the columns Sepal.Length, Sepal.Width, and Species.

```{r}
iris |> 
  dplyr::select("Sepal.Width","Petal.Length", "Species")
```

3\. Demonstrate column exclusion by selecting all columns except Sepal.Width using -Sepal.Width.

```{r}
iris |> 
  dplyr::select(-Sepal.Width)
```

4\. Contrast dplyr::select() with Base R bracket indexing by extracting rows 20 to 45 and columns 1 and 2 using iris[20:45, c(1, 2)].

```{r}
iris |> 
  slice(20:45) |>
  dplyr::select(,c(1, 2))
```

5\. Filter the dataset to extract: ○ Observations where Sepal.Length \> 6.0. ○ Observations where the species is either "virginica" OR "versicolor" using the logical OR operator (\|).

```{r}
iris |>
  dplyr::filter(Sepal.Length > 6 ,Species == c("virginica","versicolor"))
```

6.  Sort the entire dataset by Sepal.Length in descending order using arrange(desc(...)) and display the first 6 rows.

```{r}
head(arrange(iris , desc(Sepal.Length)))
```

Task 2: Feature Engineering, Conditional Mutation & Grouped Summaries Objective: Derive calculated columns using mutate(), implement conditional labels with ifelse(), and compute grouped statistics using group_by() and summarise(). Instructions:

1\. Use mutate() to create a new column Sepal.ratio defined as $\frac{\text{Sepal.Length}}{\text{Sepal.Width}}$.

```{r}
iris |>
  mutate(
    sepal.ratio = Sepal.Length/Sepal.Width
  )
```

2\. Chain mutate() with select() using the pipe (\|\>) to produce a two-column summary data frame containing only Species and Sepal.ratio.

```{r}
iris |> 
  mutate(sepal.ratio = Sepal.Length / Sepal.Width) |>
  select(Species, sepal.ratio)
```

3\. Create a derived categorical column named Petal.Size using mutate() and ifelse() such that: ○ If Petal.Length \> 4, assign "Large". ○ Otherwise, assign "Small".

```{r}
iris |>
  mutate(
    Petal.size = ifelse(Petal.Length > 4, "Large", "Small")
  )
```

4.  Calculate overall summary statistics for Sepal.Length: its mean (mean_sepal), minimum (min_sepal), and maximum (max_sepal).

    ```{r}
    iris |>
      summarise(
        mean_sepal = mean(Sepal.Length),
        min_sepal = min(Sepal.Length),
        max_sepal = max(Sepal.Length)
      )
    ```

5.  Group the data by Species using group_by() and compute: ○ Mean of Sepal.Length (mean_sepal) ○ Standard deviation of Sepal.Length (sd_sepal) ○ Total number of observations per species (count = n())

```{{r}}
iris |>
  group_by(Species) |>
  summarise(
    mean_sepal = mean(Sepal.Length),
    sd_sepal = sd(Sepal.Length),
    count = n()
  )
```

6.  Build a grouped pipeline that calculates the average petal length per species (average_petal = mean(Petal.Length)) and filters for species where average_petal \> 4.

```{{r}}
iris |>
  group_by(Species)|>
  summarise(
    average_petal = mean(Petal.Length))|>
  filter(average_petal>4)
```

Task 3: Missing Value Handling & Data Reshaping (tidyr) Objective: Identify, drop, and replace missing values (NA), and pivot datasets between wide and long representations.

Instructions: 1. Construct the following test data frame named DF: R DF \<- data.frame( name = c("A", "B", "C", "D"), salary = c(50000, NA, 60000, NA), age = c(25, 30, NA, 35) )

```{r}
DF = data.frame(
  name = c("A","B","C","D"),
  salary = c(50000, NA, 60000, NA),
  age = c(25, 30, NA, 35)
)
```

2\. Check for missing values across the entire data frame using is.na(DF).

```{r}
is.na(DF)
```

3\. Drop all rows containing any NA value using drop_na().

```{r}
library("tidyr")
DF |>
  drop_na()
```

4\. Replace missing values using replace_na() by providing a named list: impute salary = 0 and age = 0.

```{r}
DF %>%
  replace_na(
    list(
      salary=0,
      age=0
    ))
```

5\. Construct the student marks data frame: R Marks \<- data.frame( Student = c("A", "B", "C"), Maths = c(80, 70, 90), Science = c(75, 85, 88), English = c(90, 80, 85) )

```{r}
marks <-
  data.frame(
    Student = c("A","B","C"),
    Maths = c(80,70,90),
    Science = c(75,85,88),
    English = c(90,80,85)
  )
```

6\. Pivot Marks from wide to long format using pivot_longer(): ○ Gather columns Maths:English. ○ Name the new key column "Subject". ○ Name the new value column "Marks". ○ Store the result in long_marks.

```{r}
long_marks = marks |>
  pivot_longer(
    cols = Maths:English,
    names_to = "Subject",
    values_to = "Marks"
  )
```

7\. Pivot long_marks back to wide format using pivot_wider() and verify that the output matches the original Marks table.

```{r}
wide_marks = long_marks |>
  pivot_wider(
    names_from = Subject,
    values_from = Marks
  )
```

Task 4: Relational Two-Table Operations (Database Joins) Objective: Merge relational tables using the four fundamental join types: Inner, Left, Right, and Full joins.

Instructions: 1. Construct two relational tables: R Students \<- data.frame( ID = c(1, 2, 3, 4), Name = c("A", "B", "C", "D") ) marks \<- data.frame( ID = c(1, 2, 3, 5), Marks = c(80, 75, 90, 60) )

```{r}
students <- data.frame(
  ID=c(1,2,3,4),
  Names=c("A","B","c","D")
)

marks <- data.frame(
  ID = c(1,2,3,5),
  Marks = c(80,75,90,60)
)
```

2.  Execute and print the outputs of all four relational joins using the key column by = "ID": ○ inner_join(Students, marks, by = "ID") ○ left_join(Students, marks, by = "ID") ○ right_join(Students, marks, by = "ID") ○ full_join(Students, marks, by = "ID")
3.  Write a brief comment above each join explaining: ○ Which rows were retained. ○ Which rows were dropped. ○ Where and why NA values were introduced.

```{r}
#1)inner_join(): Returns only rows where there is a match in both data frames, discarding unmatched rows from either side. 
inner_join(students, marks, by = "ID")

#2)left_join(): Returns all rows from the first data frame (x) and matching rows from the second (y), filling in NA for non-matches in y. 
left_join(students, marks, by = "ID")

#3)right_join(): Returns all rows from the second data frame (y) and matching rows from the first (x), filling in NA for non-matches in x. 
right_join(students, marks, by = "ID")

#4)full_join(): Returns all rows from both data frames, matching where possible and using NA for missing values on either side.
full_join(students, marks, by = "ID")
```

Task 5: Visual Analytics with ggplot2 Objective: Construct graphics using the Grammar of Graphics (ggplot2), demonstrating points, lines, bars, histograms, and density curves.

Instructions: 1. Load library(ggplot2) and set up the default iris dataset.

```{r}
library(ggplot2)
data(iris)
```

2\. Scatter Plot (geom_point): Plot x = Sepal.Length vs. y = Sepal.Width, map point size to Petal.Length, and map point color to Species. Add clear title and axis labels.

```{r}
ggplot(iris,
       aes(#asthetic
         x = Sepal.Length,
         y = Sepal.Width, 
         size = Petal.Length,
         color = Species)
) + geom_point()
```

3\. Line Plot (geom_line): Plot x = Sepal.Width vs. y = Petal.Length.

```{r}
ggplot(iris,
       aes(
         x = Sepal.Width,
         y = Petal.Length
       )
) + geom_line()
```

4\. Bar Plot (geom_bar): Plot the distribution of Species, mapping fill color to Species.

```{r}
ggplot(iris,
       aes(x= Species,
           fill = Species)
) + geom_bar()
```

5\. Histogram (geom_histogram): Plot the distribution of Sepal.Length using bins = 5, fill = "skyblue", and colour = "black".

```{r}
ggplot(iris,
       aes(x = Sepal.Length))+ 
  geom_histogram(bins =5,
                 fill = "skyblue",
                 colour = "black")
```

6\. Density Plot (geom_density): Plot smoothed density estimates of Sepal.Length grouped by Species, using alpha = 0.8 for fill transparency.

```{r}
ggplot(
  iris,
  aes(
    x = Sepal.Length,
    fill = Species
  )
) + geom_density(alpha = 0.4)
```
