---
title: "63_Lab4_Visualization"
output: html_document
---

```{r}
rm(list = ls())
```

Task 1: Time-Series & Multi-Panel Layouts (Gasoline prices.csv) Objective: Parse dates, aggregate time-series metrics into quarterly intervals using the zoo package, and build multi-paneled graphical dashboards. Instructions:

1\. Load Gasoline prices.csv and inspect its structure using str() and head().

```{r}
library(readr)
Gasoline_prices <- read_csv("D:/Uni/SY/Sem-3/R-Programming/employee/Gasoline prices.csv")
str(Gasoline_prices)
head(Gasoline_prices)
```

2.  Convert the Date column from character/factor into a Date object using as.Date() with the correct format parameter ("%m-%d-%Y").

```{r}
weekly_gasoline_prices = Gasoline_prices

weekly_gasoline_prices$Date = as.Date(weekly_gasoline_prices$Date, format = "%m-%d-%Y")

str(weekly_gasoline_prices$Date)   
```

3\. Using library(zoo), construct a zoo time-series object named wpg indexed by the formatted Date column.

```{r}
library(zoo)
wpg <- zoo(weekly_gasoline_prices$Dollars.per.Gallon, order.by = weekly_gasoline_prices$Date)
```

4.  Use aggregate(wpg, as.yearqtr) to aggregate the prices into quarterly averages stored in wpg1.

```{r}
wpg1 <- aggregate(wpg, as.yearqtr, mean)
```

5\. Configure a $2 \times 2$ grid layout using par(mfrow = c(2, 2), mar = c(2, 3, 2, 1)). 6. Render the following four sub-plots in the grid: ○ Plot 1: Line plot (type = "l") of wpg1 with col = "red" and lwd = 2.

```{r}

```

○ Plot 2: Line plot (type = "l") of raw weekly_gasoline_prices with col = "green" and lwd = 2.

```{r}

```

○ Plot 3: Bar plot (barplot()) of wpg1 with col = "blue" and a border = "red".

```{r}

```

○ Plot 4: Standard point-line plot of wpg1 with customized title and labeled axes.

```{r}

```

Task 2: Categorical Distribution & Legends (nations_dataset.csv) Objective: Clean factor structures, construct frequency tables, and build customized bar plots with color palettes and legends. Instructions: 1. Load nations_dataset.csv and inspect the variable data types using str().

```{r}
library(readr)
nations_dataset <- read_csv("D:/Uni/SY/Sem-3/R-Programming/employee/nations_dataset.csv")
str(nations_dataset)
```

2\. Convert the outlook column to a nominal factor using as.factor().

```{r}

```

3.  Compute the frequency distribution of outlook using table().

```{r}

```

4\. Plot a bar chart of the frequency table with: ○ Custom title: "Economic Outlook Frequency Distribution". ○ Rainbow palette containing distinct colors matching the number of unique factor levels. ○ Labeled X-axis ("Outlook Category") and Y-axis ("Number of Nations").

```{r}

```

5\. Add a legend to the top-right corner showing each category label, matching fill colors, and an adjusted scale factor (cex = 0.7).

```{r}

```

Task 3: Relative Frequencies, 2D & 3D Pie Charts (softdrinks.xlsx) Objective: Calculate relative percentages, build annotated 2D pie charts, and render exploded 3D pie charts using plotrix. Instructions: 1. Read softdrinks.xlsx using library("readxl"). Convert the drinknames column into a factor.

```{r}

```

2.  Build a frequency table and convert it to a data frame named softdrink1 using cbind.data.frame(table(softdrinks)).

    ```{r}

    ```

3.  Create a derived column named relativefreq calculated as: $$\text{relativefreq} = \frac{\text{Freq}}{\sum \text{Freq}}$$

    ```{r}

    ```

4.  Generate horizontal and vertical bar plots of relativefreq with category names displayed along the axes.

    ```{r}

    ```

5.  Create a 2D pie() chart where each slice is labeled with the brand name and its percentage value formatted using paste0() and round() (e.g., "Coca-Cola 20.5%").

    ```{r}

    ```

6.  Using library("plotrix"), generate a 3D pie chart (pie3D()) with the same percentage labels, matching color palettes, and a customized top-right legend.

    ```{r}

    ```

    Task 4: Histograms, Frequency Labels & Dual-Axis Density Plots Objective: Understand bin widths (breaks), overlay frequency labels on bars, and construct dual-axis plots combining histograms with kernel density estimates. Instructions

    1.  Generate a reproducible sample of 1,000 random integers between 50 and 100 using sample(50:100, 1000, replace = TRUE).

        ```{r}

        ```

    2.Plot three baseline histograms comparing the effect of different bin counts: breaks = 100, breaks = 10, and breaks = 1.

    ```{r}

    ```

    3.  Save the histogram object into a variable m \<- hist(...) using breaks = 10.

        ```{r}

        ```

    4.Use text() with m$mids and m$counts to add exact frequency counts directly above each histogram bar.

    ```{r}

    ```

    5.Configure margins using par(mar = c(5, 4, 4, 6) + 0.1). Plot the histogram, set par(new = TRUE), and overlay a kernel density curve (lines(density(var)) or plot(density(var))) with lwd = 3.

    ```{r}

    ```

    6.Add a secondary Y-axis on the right-hand side (axis(4)) labeled "Density Estimate" using mtext(side = 4).

    ```{r}

    ```

    Task 5: Stacked and Grouped Multi-Variable Matrix Bar Plots (Meal.xlsx) Objective: Subset cross-tabulated survey responses, cast data frames into matrices, and build side-by-side vs. stacked visual comparisons. Instructions:

    1.Import Meal.xlsx using read_excel().

    ```{r}

    ```

    2.Extract the price evaluation columns across the top 3 quality ratings (Meal[c(1:3), c(2, 6)] or matching numeric categories).

    ```{r}

    ```

    3.Convert this subset to a numeric matrix using as.matrix().

    ```{r}

    ```

    4.Plot a Stacked Bar Plot of the matrix with distinct colors per category, white borders, labeled axes, and an associated legend showing the quality ratings.

    ```{r}

    ```

    5.Plot a Grouped (Side-by-Side) Horizontal Bar Plot by setting beside = TRUE and horiz = TRUE. Add appropriate axis labels and legends.

    ```{r}

    ```
