---
title: "63_Lab3_Visualization"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r}
rm(list = ls())
```

Task 1: Time-Series & Multi-Panel Layouts (Gasoline prices.csv) Objective: Parse dates, aggregate time-series metrics into quarterly intervals using the zoo package, and build multi-paneled graphical dashboards.
Instructions:

1\.
Load Gasoline prices.csv and inspect its structure using str() and head().

```{r}
library(readr)
Gasoline_prices <- read_csv("D:/Uni/SY/Sem-3/R-Programming/employee/Gasoline prices.csv")
str(Gasoline_prices)
head(Gasoline_prices)
```

2.
Convert the Date column from character/factor into a Date object using as.Date() with the correct format parameter ("%m-%d-%Y").

```{r}
weekly_gasoline_prices = Gasoline_prices

weekly_gasoline_prices$Date = as.Date(weekly_gasoline_prices$Date, format = "%m-%d-%Y")

str(weekly_gasoline_prices$Date) 
```

3\.
Using library(zoo), construct a zoo time-series object named wpg indexed by the formatted Date column.

```{r}
library(zoo)

wpg <- zoo(weekly_gasoline_prices[[2]], order.by = weekly_gasoline_prices$Date)   
```

4\.
Use aggregate(wpg, as.yearqtr) to aggregate the prices into quarterly averages stored in wpg1.

```{r}
wpg1 <- aggregate(wpg, as.yearqtr, mean)
```

5\.
Configure a $2 \times 2$ grid layout using par(mfrow = c(2, 2), mar = c(2, 3, 2, 1)).

```{r}
par(mfrow = c(2, 2), mar = c(2, 3, 2, 1))
```

6\.
Render the following four sub-plots in the grid: ○ Plot 1: Line plot (type = "l") of wpg1 with col = "red" and lwd = 2.

```{r}
plot(wpg1, type = "l", col = "red", lwd = 2)
```

○ Plot 2: Line plot (type = "l") of raw weekly_gasoline_prices with col = "green" and lwd = 2.

```{r}
plot(weekly_gasoline_prices, type = "l", col = "green", lwd = 2)
```

○ Plot 3: Bar plot (barplot()) of wpg1 with col = "blue" and a border = "red".

```{r}
barplot(wpg1, col = "blue", border = "red")
```

○ Plot 4: Standard point-line plot of wpg1 with customized title and labeled axes.

```{r}
plot(wpg1, type = "b", 
     main = "Weekly Gasoline Prices", 
     xlab = "Week", 
     ylab = "Price")
```

Task 2: Categorical Distribution & Legends (nations_dataset.csv) Objective: Clean factor structures, construct frequency tables, and build customized bar plots with color palettes and legends.
Instructions:

1\.
Load nations_dataset.csv and inspect the variable data types using str().

```{r}
Nations <- read_csv("D:/Uni/SY/Sem-3/R-Programming/employee/nations_dataset.csv")
str(Nations)
```

2\.
Convert the outlook column to a nominal factor using as.factor().

```{r}
Nations$outlook <- as.factor(Nations$outlook)
```

3\.
Compute the frequency distribution of outlook using table().

```{r}
table(Nations$outlook)
```

4.
Plot a bar chart of the frequency table with: ○ Custom title: "Economic Outlook Frequency Distribution".
○ Rainbow palette containing distinct colors matching the number of unique factor levels.
○ Labeled X-axis ("Outlook Category") and Y-axis ("Number of Nations").

5\.
Add a legend to the top-right corner showing each category label, matching fill colors, and an adjusted scale factor (cex = 0.7).

```{r}
barplot(table(Nations$outlook), main="Bar Plot", horiz = T, col = rainbow(3),ylab = "Outlook Category", xlab = "Number of Nations")

legend("topright", c("Negative", "Positive", "Stable"), fill = rainbow(3), cex = 0.7)
```

Task 3: Relative Frequencies, 2D & 3D Pie Charts (softdrinks.xlsx) Objective: Calculate relative percentages, build annotated 2D pie charts, and render exploded 3D pie charts using plotrix.
Instructions:

1\.
Read softdrinks.xlsx using library("readxl").
Convert the drinknames column into a factor.

```{r}
library(readr)

softdrinks <- read_csv("D:/Uni/SY/Sem-3/R-Programming/employee/softdrinks.csv")

softdrinks$drinknames = as.factor(softdrinks$drinknames)
```

2.  Build a frequency table and convert it to a data frame named softdrink1 using cbind.data.frame(table(softdrinks)).

    ```{r}
    softdrink1 = cbind.data.frame(table(softdrinks))
    ```

3.  Create a derived column named relativefreq calculated as: $$\text{relativefreq} = \frac{\text{Freq}}{\sum \text{Freq}}$$

    ```{r}
    softdrink1$relativefreq = softdrink1$Freq/sum(softdrink1$Freq)
    ```

4.  Generate horizontal and vertical bar plots of relativefreq with category names displayed along the axes.

    ```{r}
    barplot(softdrink1$relativefreq, names.arg = softdrink1$drinknames, col= rainbow(8))
    ```

5.  Create a 2D pie() chart where each slice is labeled with the brand name and its percentage value formatted using paste0() and round() (e.g., "Coca-Cola 20.5%").

    ```{r}
    library("plotrix")
    pie(
      softdrink1$relativefreq, 
      labels = paste0(softdrink1$drinknames, " ", round(100 * softdrink1$relativefreq, 1), "%"), 
      main = "Soft Drinks Frequency Distribution", 
      col = rainbow(length(softdrink1$drinknames))
    )
    ```

6.  Using library("plotrix"), generate a 3D pie chart (pie3D()) with the same percentage labels, matching color palettes, and a customized top-right legend.

```{r}
library("plotrix")
pie3D(softdrink1$relativefreq, 
      labels = paste0(softdrink1$drinknames, " " ,round(100*softdrink1$reletivefreq, 2),"%"),
      main = "soft drinks frequency dust",
      col = rainbow(8))
```

Task 4: Histograms, Frequency Labels & Dual-Axis Density Plots Objective: Understand bin widths (breaks), overlay frequency labels on bars, and construct dual-axis plots combining histograms with kernel density estimates.
Instructions:

1\.
Generate a reproducible sample of 1,000 random integers between 50 and 100 using sample(50:100, 1000, replace = TRUE).

```{r}
var = sample(50:100, 1000, replace = T)
summary(var)
```

2.
Plot three baseline histograms comparing the effect of different bin counts: breaks = 100, breaks = 10, and breaks = 1.

```{r}
hist(var, xlab="freq", ylab="varaible", col=rainbow(20), breaks = 100)
hist(var, xlab="freq", ylab="varaible", col=rainbow(20), breaks = 10)
hist(var, xlab="freq", ylab="varaible", col=rainbow(20), breaks = 1)
```

3.
Save the histogram object into a variable m \<- hist(...) using breaks = 10.

4\.
Use text() with m$mids and m$counts to add exact frequency counts directly above each histogram bar.

```{r}
m = hist(var, xlab="freq", ylab="varaible", col=rainbow(20), breaks = 10, main = "Histogram")

text(m$mids, m$counts, labels = m$counts, adj = c(0.5, -0.5))
```

5\.
Configure margins using par(mar = c(5, 4, 4, 6) + 0.1).
Plot the histogram, set par(new = TRUE), and overlay a kernel density curve (lines(density(var)) or plot(density(var))) with lwd = 3.

```{r}
par(mar = c(5,4,4,6+0.1))
```

6\.
Add a secondary Y-axis on the right-hand side (axis(4)) labeled "Density Estimate" using mtext(side = 4).

```{r}
plot(density(var), lwd = 4, col="black", xlab = "", ylab = "0", axes = F, main = "")

axis(4, col = "black", color.axis = "black")
mtext("Cell density",side=4,col="black", line=4)
```

Task 5: Stacked and Grouped Multi-Variable Matrix Bar Plots (Meal.xlsx) Objective: Subset cross-tabulated survey responses, cast data frames into matrices, and build side-by-side vs. stacked visual comparisons.
Instructions:

1\.
Import Meal.xlsx using read_excel().

```{r}
library(readxl)
Meal <- read_excel("D:/Uni/SY/Sem-3/R-Programming/employee/Meal.xlsx")
```

2\.
Extract the price evaluation columns across the top 3 quality ratings (Meal[c(1:3), c(2, 6)] or matching numeric categories).

```{r}
Meal[c(1:3),c(2,6)]
```

3\.
Convert this subset to a numeric matrix using as.matrix().

```{r}
as.matrix(Meal[c(1:3),c(2,6)])
```

4\.
Plot a Stacked Bar Plot of the matrix with distinct colors per category, white borders, labeled axes, and an associated legend showing the quality ratings.

5\.
Plot a Grouped (Side-by-Side) Horizontal Bar Plot by setting beside = TRUE and horiz = TRUE.
Add appropriate axis labels and legends.

```{r}
barplot(as.matrix(Meal[c(1:3),c(2,6)]), col = rainbow(3), border = "white", xlab = "Price", ylab = "Number of Rest", beside = T, horiz = T)

legend("topright", Meal$`Quality rating` [1:3], cex=1, fill=rainbow(3))
```
